\chapter{Data Types}
\label{ch2}
\index{data types}

	$\Psi$lab is equipped to recognize several primitive data types.
 Scalar objects are constants, booleans, polynomials, strings and 
rationals (quotients of polynomials). These objects in turn allow to 
define matrices which admits these scalars as entries.
Other basic objects are lists and macros.
%
The objective of this chapter is to describe the use of each of 
these data types.

\section{Special Constants}
\label{s2.1}
\index{data types!constants}
\index{constants}

	$\Psi$lab provides special constants {\tt \%i}, {\tt \%pi},
{\tt \%e}, and {\tt \%eps} as primitives.  The {\tt \%i} constant
represents the imaginary number $i=\sqrt{-1}$, {\tt \%pi} is the
constant $\pi=3.1415927\cdots$ , {\tt \%e}
is the trigonometric constant $e=2.7182818\cdots$, and {\tt \%eps}
is a constant representing the precision of the machine ({\tt \%eps}
is the biggest number for which $1+\mbox{\tt \%eps}=1$). {\tt \%inf}
and {\tt \%nan} stand for ``infinity'' and ``NotANumber'' respectively.

Finally boolean constants are {\tt \%t} and {\tt \%f} which stand for
''true'' and ''false'' respectively. Note that {\tt \%t} is the
same as {\tt 1==1} and {\tt \%f} is the same as {\verb!~%t!}.

\section{Constant Matrices}
\label{s2.2}
\index{data types!matrices}

	$\Psi$lab labels a number of data objects as matrices.  
Scalars, vectors, and matrices whose entries are either real or complex
are all considered as matrices.  The details of the use
of these objects are revealed in the following $\Psi$lab sessions.

\paragraph{Scalars}\index{scalars}
Scalars are either real or complex numbers.  The values of
scalars can be assigned to variable names chosen by the user.
\begin{verbatim}
--> a=5+2*%i
 a         =
 
    5. + 2.i  
 
--> B=-2+%i;
 
--> a*b
 ans       =
 
  - 12. + i    
 
--> c=a+b;
 
--> c
 c         =
 
    3. + 3.i  
\end{verbatim}
Notice that $\Psi$lab evaluates immediately lines that
end with a carriage return.  Instructions that end in a semi-colon
are evaluated but are not displayed at the terminal. Lower-case
and Upper-case letters are equivalent, except in character strings.

\paragraph{Vectors}\index{vectors}
The usual way of creating vectors is with brackets
\begin{verbatim}
 
--> v=[2 -3+%i 7]
 v         =
 
!   2.  - 3. + i      7. !
 
--> v'
 ans       =
 
!   2.       !
! - 3. - i   !
!   7.       !
 
--> w=[-3;-3-%i;2]
 w         =
 
! - 3.       !
! - 3. - i   !
!   2.       !
 
--> v'+w
 ans       =
 
! - 1.       !
! - 6. - 2.i !
!   9.       !
 
--> v*w
 ans       =
 
    18.  
 
--> w'.*v
 ans       =
 
! - 6.    8. - 6.i    14. !
\end{verbatim}
Notice that vector elements that are separated by blanks (or by commas)
yield row vectors and those separated by semi-colons give column vectors.
Use a single quote for transposing a vector\index{vectors!transpose} 
(noting that 
one obtains the complex conjugate for complex entries).  Vectors of like
dimension can be added and subtracted.  The scalar product of a row and
column vector is demonstrated above.  Element by element
multiplication ({\tt .*}) and division ({\tt ./}) is also possible 
as was demonstrated.

	Vectors\index{vectors!incremental} of elements which increase
or descend in increments are easily made
\begin{verbatim}
 
--> v=5:-.5:3
 v         =
 
!   5.    4.5    4.    3.5    3. !
\end{verbatim}
The created vector begins with the first value and ends
with the third value stepping in increments of the second value.
When not specified the default increment is unity.  A constant vector
can be created using the {\tt ones}\index{ones@{\tt ones}}\index{vectors!constant} 
facility 
\begin{verbatim}
 
--> v=[1 5 6]
 v         =
 
!   1.    5.    6. !
 
--> ones(v)
 ans       =
 
!   1.    1.    1. !
 
--> ones(v')
 ans       =
 
!   1. !
!   1. !
!   1. !
 
--> ones(1:4)
 ans       =
 
!   1.    1.    1.    1. !
 
--> 3*ones(1:4)
 ans       =
 
!   3.    3.    3.    3. !
\end{verbatim}
Notice that {\tt ones} replaces its vector argument by a vector 
of equivalent dimensions filled with ones.

\paragraph{Matrices}\index{matrices}\index{data types!matrices}
Row elements are separated by spaces
or commas and column elements by semi-colons.  Multiplication
of matrices by scalars, vectors, or other matrices is in the usual
sense.  Addition and
subtraction of matrices is element by element and element by element
multiplication and division can be accomplished with the {\tt .*}
and {\tt ./} operators.
\begin{verbatim}
 
--> a=[2 1 4;5 -8 2]
 a         =
 
!   2.    1.    4. !
!   5.  - 8.    2. !
 
--> b=ones(2,3)
 b         =
 
!   1.    1.    1. !
!   1.    1.    1. !
 
--> a.*b
 ans       =
 
!   2.    1.    4. !
!   5.  - 8.    2. !
 
--> a*b'
 ans       =
 
!   7.    7. !
! - 1.  - 1. !
\end{verbatim}
Notice that the {\tt ones}\index{ones@{\tt ones}}\index{matrices!constant}
operator with two arguments separated
by a comma creates a matrix of ones using the the arguments as
dimensions.
Matrices can be used as elements to larger 
matrices\index{matrices!block construction}.  Furthermore,
the dimensions of a matrix can be changed.
\begin{verbatim}
 
--> a=[1 2;3 4]
 a         =
 
!   1.    2. !
!   3.    4. !
 
--> b=[5 6;7 8]
 b         =
 
!   5.    6. !
!   7.    8. !
 
--> c=[9 10;11 12]
 c         =
 
!   9.     10. !
!   11.    12. !
 
--> d=[a,b,c]
 d         =
 
!   1.    2.    5.    6.    9.     10. !
!   3.    4.    7.    8.    11.    12. !
 
--> e=matrix(d,3,4)
 e         =
 
!   1.    4.    6.    11. !
!   3.    5.    8.    10. !
!   2.    7.    9.    12. !
\end{verbatim}
Notice that the matrix {\tt d} is created by using other matrix
elements.  The {\tt matrix}\index{matrix@{\tt matrix}} 
primitive creates a new matrix using
the dimensions specified by the second two arguments.  The element
ordering in the matrix {\tt d} is top to bottom and then left to right
which explains the ordering of the re-arranged matrix in {\tt e}.

\section{Matrices of Character Chains}
\label{s2.3}
\index{data types!character chains}
\index{character chains}

	Character chains can be created by using single quotes.
Concatenation of chains is performed by the {\tt +} operation.
Matrices of character chains are constructed as ordinary matrices,
e.g. using brackets.
A very important feature of matrices of character chains is
the capacity to manipulate and create macros.  Furthermore,
symbolic manipulation of mathematical objects can be implemented
using matrices of character chains.  The following illustrates
some of these features. 
\begin{verbatim}
 
--> x=1;y=2;z=3;w=4;v=5;
 
--> a=['x' 'y';'z' 'w+v']
 a         =
 
!x  y    !
!        !
!z  w+v  !
 
--> at=trianfml(a)
 at        =
 
!z  w+v          !
!                !
!0  z*y-x*(w+v)  !
 
--> evstr(at)
 ans       =
 
!   3.    9. !
!   0.  - 3. !
\end{verbatim}
Note that in the above $\Psi$lab session the macro 
{\tt trianfml}\index{trianfml@{\tt trianfml}}\index{symbolic triangularization}
performs the symbolic triangularization of the matrix {\tt a}.
The value of the resulting symbolic matrix can be obtained by
using {\tt evstr}.  

	A very important aspect of character chains is that they
can be used to automatically create new macros (for more on macros
see Section~\ref{s4.2}).  An example of automatically creating a 
macro is illustrated in the following $\Psi$lab session where it is
desired to study a polynomial of two variables {\tt s} and {\tt t}.
Since polynomials in two independent variables is not directly 
supported in $\Psi$lab we can construct a new data structure using
a list (see Section~\ref{s2.5}).
The polynomial to be studied is $(t^2+2t^3)-(t+t^2)s+ts^2+s^3$.
\begin{verbatim}
-->getf("../macros/make_macro.sci");
 
-->s=poly(0,'s');
 
-->t=poly(0,'t');
 
-->p=list(t^2+2*t^3,-t-t^2,t,1+0*t);
 
-->pst=makemacro(p)
 pst       =
 
[p]=pst(t)
 
-->pst
 pst       =
 
[p]=pst(t)
 
-->pst(1)
 ans       =
 
              2   3  
    3 - 2s + s + s   
\end{verbatim}
Here the polynomial is represented by the command which puts
the coefficients of the variable {\tt s} in the list {\tt p}.
The list {\tt p} is then processed by the macro {\tt makemacro}
which makes a new macro {\tt pst}.  The contents of the new macro
can be displayed and this macro can be evaluated
at values of {\tt t}.  The creation of the new macro {\tt pst}
is accomplished as follows
\begin{verbatim}
function [newmacro]=makemacro(p)
   n=size(p);
   num=mulf(makestr(p(1)),'1');
   for k=2:n,
      new=mulf(makestr(p(k)),'s^'+string(k-1));
      num=addf(num,new);
   end,
   text='p='+num;
   deff('<p>=newmacro(t)',text),


function [str]=makestr(p)
   n=degree(p)+1,
   c=coeff(p),
   str=string(c(1)),
   x=part(varn(p),1),
   xstar=x+'^',
   for k=2:n, 
      ck=c(k), 
      if ck<>0 then,
      str=addf(str,mulf(string(c(k)),(xstar+string(k-1))));
      end;
   end,

\end{verbatim}
Here the macro {\tt makemacro} takes the list {\tt p} and creates the
macro {\tt pst}.  Inside of {\tt makemacro} there is a call to another 
macro {\tt makestr} which makes the string which represents each 
term of the new two variable polynomial.  The macros {\tt addf} and
{\tt mulf} are for adding and multiplying strings (i.e., 
{\tt addf(x,y)} yields the string {\tt x+y}).  Finally, the 
essential command for creating the new macro 
is the primitive {\tt deff}.  The {\tt deff} primitive 
creates a macro defined by two matrices
of character strings.  Here the 
macro {\tt p} is defined by the two character strings
{\tt '[p]=newmacro(t)'} and {\tt text} where the string {\tt text}
evaluates to the polynomial in two variables.

\section{Polynomials and Matrix Polynomials}
\label{s2.4}
\index{data types!polynomials}
\index{polynomials}

	Polynomials are easily created and manipulated in $\Psi$lab.
Essentially the manipulation of polynomial matrices is identical
to that for constant matrices.
The {\tt poly}\index{poly@{\tt poly}} 
primitive in $\Psi$lab can be used to specify the coefficients
of a polynomial or the roots of a polynomial
\begin{verbatim}
--> p=poly([1 2],'s')
 p         =
 
              2  
    2 - 3s + s   
 
--> q=poly([1 2],'s','c')
 q         =
 
    1 + 2s   
 
--> p+q
 ans       =
 
             2  
    3 - s + s   
 
--> p*q
 ans       =
 
              2    3  
    2 + s - 5s + 2s   
 
--> q/p
 ans       =
 
      1 + 2s     
    -----------  
              2  
    2 - 3s + s   
\end{verbatim}
Notice that the polynomial {\tt p} has the {\em roots}
1 and 2 whereas the polynomial {\tt q} has the {\em coefficients}
1 and 2.  It is the third argument in the {\tt poly} primitive which
specifies the coefficient flag option.  
In the case where the first argument of {\tt poly} is a square matrix
and the roots option is in effect the result is the characteristic
polynomial of the matrix.
\begin{verbatim}
 
--> poly([1 2;3 4],'s')
 ans       =
 
              2  
  - 2 - 5s + s   
\end{verbatim}
Polynomials can be added,
subtracted, multiplied, and divided, as usual, but only between polynomials
of like formal variable.

	Polynomials, like real and complex constants, can be used
as elements in matrices.  This is a very useful feature of $\Psi$lab
for systems theory.
\begin{verbatim}
 
--> s=poly(0,'s')
 s         =
 
    s   
 
--> a=[1 s;s 1+s^2]
 a         =
 
!   1     s      !
!                !
!              2 !
!   s     1 + s  !
 
--> b=[1/s 1/(1+s);1/(1+s) 1/s^2]
 b         =
 
!   1           1    !
! ------      ------ !
!   s         1 + s  !
!                    !
!     1       1      !
!    ---     ---     !
!              2     !
!   1 + s     s      !

\end{verbatim}
From the above examples it can be seen that matrices can be constructed
of polynomials and rational polynomials.

\section{Boolean matrices}
\index{data types!booleans}
\index{boolean}

Boolean constants are {\tt \%t} and {\tt \%f}. They can be used in
boolean matrices. The syntax is the same as for ordinary matrices i.e.
they can be concatenated, transposed, etc...

Operations symbols used with boolean matrices or used to create
boolean matrices are {\tt ==}, {\tt ~}.

If {\tt B} is a matrix of booleans {\tt or(B)} and {\tt and(B)} 
perform the logical {\tt or} and {\tt and}.

\section{Lists, Linear systems}
\label{s2.5}
\index{data types!lists}
\index{lists}

	$\Psi$lab has a list data type.  The list is a collection of data
objects not necessarily of the same type.  A list can contain any of
the already discussed data types as well as other lists, macros, and
libraries.  Lists are useful for defining structured data objects.
For example, in $\Psi$lab linear systems are treated as lists.
The basic macro which is used for defining linear systems is {\tt syslin}.
This macro receives as parameters the constant matrices which
define a linear system in state-space form or, in the case of
system in transfer form its input must be a rational matrix.
To be more specific, the calling sequence of {\tt syslin} is
either {\tt Sl=syslin('dom',A,B,C,D,x0)} or {\tt Sl=syslin('dom',trmat)}.
{\tt dom} is one of the character strings {\tt 'c'} or {\tt 'd'}
for continuous time or discrete time systems respectively.
It is useful to note that {\tt D} can be a polynomial matrix 
(improper systems). ({\tt D} and {\tt x0} are optional arguments).
{\tt trmat} is a rational matrix i.e. it is defined as a matrix
of rationals (ratios of polynomials).
Conversion from a representation to another is done by {\tt ss2tf}
or {\tt tf2ss}. Improper systems are also treated.

\begin{verbatim}
 
--> A=[0 -1;1 -3];B=[0;1];C=[-1 0];
 
--> h=syslin('c',A,B,C)
 h         =
 
 
       h(1)   (state-space system:)
 
 lss   
 
       h(2) = A matrix = 
 
!   0.  - 1. !
!   1.  - 3. !
 
       h(3) = B matrix = 
 
!   0. !
!   1. !
 
       h(4) = C matrix = 
 
! - 1.    0. !
 
       h(5) = D matrix = 
 
    0.  
 
       h(6) = X0 (initial state) = 
 
!   0. !
!   0. !
 
       h(7) = Time domain = 
 
 c   
 
--> hs=ss2tf(h)
 hs        =
 
        1        
    ----------   
              2  
    1 + 3s + s   
 
--> size(hs)
 ans       =
 
!   1.    1. !
 
--> hs(1)
 ans       =
 
 r   
 
--> hs(2)
 ans       =
 
    1   
 
--> hs(3)
 ans       =
 
              2  
    1 + 3s + s   
 
--> hs(4)
 ans       =
 
 c   
 
--> typeof(hs)
 ans       =
 
    rational

-->inv(hs)         //inversion of transfer matrix
 ans       =
 
              2  
    1 + 3s + s   
    -----------  
        1        
 
-->ss2tf(inv(h))    //inversion in state-space
 ans       =
 
              2  
    1 + 3s + s   
    -----------  
        1        
\end{verbatim}
As can be seen by the above $\Psi$lab session the list {\tt h} begins
with the character string {\tt 'lss'} which in this case 
indicates that the list represents a linear system.
The five ensuing list elements are matrices which give the
state space description of the linear system and its initial condition
($\dot{x}=Ax+Bu$, $y=Cx+Du$, $x(0)=x_0$).  Finally, the list element
{\tt 'c'} indicates that the list represents a continuous linear system.

	The list representation allows handling of the linear system as an
abstract data object.  For example, the linear system can be combined
with other linear systems or the transfer function representation of
the linear system can be obtained as was done above using {\tt ss2tf}.
Note that the transfer function representation of the linear system
is itself a list. The list consists of four elements: the first element 
is the character string {\tt 'r'} which indicates
that the list represents a rational polynomial matrix, the 
second and third elements are the numerator and
denominator polynomials, and finally, the fourth element is the character
string {\tt 'c'} which indicates that the transfer function is that of
a continuous system.
	A very useful aspect of
the manipulation of systems in $\Psi$lab  is that a system can
be handled as a data object.
Linear systems can be 
inter-connected\index{linear systems!inter-connection of},
their representation
can be easily changed from state-space to transfer function
and vice versa.

	The inter-connection of linear systems can be made
as illustrated in Figure~\ref{f3.2}.
%
\begin{figure}[tb]
\center{
\begin{picture}(200,300)

\put(180,270){\framebox(0,0){\tt s1*s2}}
\put(9,270){\circle{2}}
\put(10,270){\vector(1,0){20}}
\put(30,260){\framebox(30,20){$s_1$}}
\put(60,270){\vector(1,0){30}}
\put(90,260){\framebox(30,20){$s_2$}}
\put(120,270){\vector(1,0){20}}
\put(141,270){\circle{2}}

\put(180,220){\framebox(0,0){\tt s1+s2}}
\put(29,220){\circle{2}}
\put(30,220){\line(1,0){20}}
\put(50,220){\circle*{2}}
\put(50,200){\line(0,1){40}}
\put(50,200){\vector(1,0){20}}
\put(50,240){\vector(1,0){20}}
\put(70,190){\framebox(30,20){$s_2$}}
\put(70,230){\framebox(30,20){$s_1$}}
\put(100,200){\line(1,0){20}}
\put(100,240){\line(1,0){20}}
\put(120,240){\vector(0,-1){15}}
\put(120,200){\vector(0,1){15}}
\put(120,220){\circle{10}}
\put(120,220){\framebox(0,0){$+$}}
\put(125,220){\vector(1,0){15}}
\put(141,220){\circle{2}}

\put(180,140){\framebox(0,0){\tt [s1,s2]}}
\put(49,160){\circle{2}}
\put(49,120){\circle{2}}
\put(50,120){\vector(1,0){20}}
\put(50,160){\vector(1,0){20}}
\put(70,110){\framebox(30,20){$s_2$}}
\put(70,150){\framebox(30,20){$s_1$}}
\put(100,120){\line(1,0){20}}
\put(100,160){\line(1,0){20}}
\put(120,160){\vector(0,-1){15}}
\put(120,120){\vector(0,1){15}}
\put(120,140){\circle{10}}
\put(120,140){\framebox(0,0){$+$}}
\put(125,140){\vector(1,0){15}}
\put(141,140){\circle{2}}

\put(180,50){\framebox(0,0){\tt [s1;s2]}}
\put(49,50){\circle{2}}
\put(50,50){\line(1,0){20}}
\put(70,50){\circle*{2}}
\put(70,30){\line(0,1){40}}
\put(70,30){\vector(1,0){20}}
\put(70,70){\vector(1,0){20}}
\put(90,20){\framebox(30,20){$s_2$}}
\put(90,60){\framebox(30,20){$s_1$}}
\put(120,30){\vector(1,0){20}}
\put(120,70){\vector(1,0){20}}
\put(141,30){\circle{2}}
\put(141,70){\circle{2}}

\end{picture}
}
\caption{Inter-Connection of Linear Systems}
\label{f3.2}
\end{figure}
%
For each of the four possible inter-connections of two systems
{\tt s1} and {\tt s2} the command which makes the inter-connection
is shown to the right of the corresponding block diagram in
Figure~\ref{f3.2}. Feedback interconnection is performed by
\verb!s1/.s2!. 

	The representation of linear systems can be in state-space
form or in transfer function form.  These two representations can
be interchanged by using the macros 
{\tt tf2ss}\index{linear systems!{\tt tf2ss}}\index{tf2ss@{\tt tf2ss}} and 
{\tt ss2tf}\index{linear systems!{\tt ss2tf}}\index{ss2tf@{\tt ss2tf}}
which change the representations of systems from transfer function
to state-space and from state-space to transfer function, respectively.
An example of the creation, the change in representation, and the
inter-connection of linear systems is demonstrated in the following
$\Psi$lab session.
\begin{verbatim}
 
--> s=poly(0,'s')
 s         =
 
    s   
 
--> ft=1/(s-1)
 ft        =
 
      1     
    ------  
  - 1 + s   
 
--> gt=1/(s-2)
 gt        =
 
      1     
    ------  
  - 2 + s   
 
--> ft=syslin('c',ft);
 
--> gt=syslin('c',gt);
 
--> gs=tf2ss(gt);

--> ssprint(gs)
 
x = | 2 |x + | 1 |u   
 
y = | 1 |x   
 
--> hs=gs*ft;
 
--> ssprint(hs)
 
.   | 2  1 |    | 0 |    
x = | 0  1 |x + | 1 |u   
 
y = | 1  0 |x   
 
--> ht=ss2tf(hs)
 ht        =
 
        1        
    ----------   
              2  
    2 - 3s + s   
 
--> gt*ft
 ans       =
 
        1        
    ----------   
              2  
    2 - 3s + s   

\end{verbatim}
The above session is a bit long but illustrates some very important
aspects of the handling of linear systems.  First, two linear systems
are created in transfer function form using the primitive 
{\tt syslin}\index{linear systems!{\tt syslin}}\index{syslin@{\tt syslin}}.
This primitive was used to label the systems in this example 
as being continuous (as opposed to
being discrete).  The primitive {\tt tf2ss} is used to convert one of the
two transfer functions to its equivalent state-space representation
which is in list form (Note that the macro {\tt ssprint} creates a more
readable format for the state-space linear system).
The following multiplication of the two systems yields their
series inter-connection.  Notice that the inter-connection 
of the two systems is effected even though one of the systems is
in state-space form and the other is in transfer function form.
The resulting inter-connection is given in state-space form.
Finally, the primitive {\tt ss2tf} is used to convert the resulting
inter-connected systems to the equivalent transfer function representation.

\section{Macros}
\label{s2.6}
\index{data types!macros}
\index{macros}

	Macros are a very useful aspect of $\Psi$lab.  Macros
are collections of commands which are executed in a
new environment thus isolating macro variables from the original
environments variables.  Macros
can be created and executed in a number of different ways.
Furthermore, macros can pass arguments, have programming features
such as conditionals and loops, and can be recursively called.
Macros can be arguments
to other macros and can be elements in lists.  The most useful
way of creating macros is by using a text editor, however, macros
can be created directly in the $\Psi$lab environment using the 
{\tt deff}\index{deff@{\tt deff}}\index{macros!{\tt deff}} primitive.
\begin{verbatim}
 
--> deff('[x]=foo(y)','if y>0 then, x=1; else, x=-1; end')
 
--> foo(5)
 ans       =
 
    1.  
 
--> foo(-3)
 ans       =
 
  - 1.  
\end{verbatim}
Usually macros are defined in a file using an editor and loaded
into $\Psi$lab with {\tt getf('filename')} or {\tt getf('filename','c')}.
This can be done also by clicking in the {\tt File operation} button.
This latter syntax loads the macro(s) in {\tt filename} and compiles
them.
The first line of {\tt filename} must be as follows:
\begin{verbatim}
function [y1,...,yn]=macname(x1,...,xk)
\end{verbatim}
where the {\tt yi}'s are output variables and the {\tt xi}'s the
input variables.

For more on the use and creation of macros see Section~\ref{s4.2}.

\section{Libraries}
\label{s2.7}
\index{data types!libraries}
\index{libraries}

	Libraries are collections of macros which can be either loaded
into the $\Psi$lab environment automatically at the moment that 
$\Psi$lab is called or loaded at a moment desired by the user.  
Libraries are created by the {\tt lib} command. Examples of librairies
are given in the {\tt SCIDIR/macros} directory. Note that in these
directory there is a ASCII file ``names'' which contains the names
of each macro of the library, a set of {\tt .sci} files which
contain the source code of the macros and a set of {\tt .bin} files
which contained the compiled code of the macros. The Makefile invokes
{\tt scilab} for compiling the macros and generating the {\tt .bin}
files. The compiled macros of a library are automatically loaded 
into $\Psi$lab at their first call.

\section{Objects}
We conclude this chapter by noting that the function {\tt typeof}
returns the type of the various $\Psi$lab objects. The following objects
are defined:
\begin{itemize}
\item{\tt usual} For matrices with real or complex entries.
\item{\tt polynomial} For polynomial matrices: coefficients can be 
real or complex.
\item{\tt boolean} For boolean matrices.
\item{\tt character} For matrices of character strings.
\item{\tt uncompiled macro} For un-compiled macros.
\item{\tt macro} For compiled macros.
\item{\tt rational} For rational matrices (or linear systems in
transfer matrix representation ({\tt syslin} lists)
\item{\tt state-space} For linear systems in state-space 
form ({\tt syslin} lists).
\item{\tt sparse} For sparse matrices.
\item{\tt list} For ordinary lists i.e. lists which do not represent
linear systems ({\tt syslin} lists).
\item{\tt library} For library definition.
\end{itemize}