\documentstyle[12pt]{article}
\textheight=680pt \textwidth=470pt
%\topmargin=-27pt
\topmargin=-117pt
\oddsidemargin=0pt \evensidemargin=0pt \marginparwidth=60pt
\title{ACCOMPANYING FILE FOR THE DISTRIBUTION OF $\Psi$lab}
\author{ }
\date{ }
\begin{document}
\maketitle{}
\thispagestyle{empty}
\pagestyle{empty}
$\Psi$lab (hereinafter referred to as SCILAB) has been 
designed and produced  by the ``Scilab group'' mainly
composed by researchers from the META2 research project at INRIA 
``Institut National de Recherche en Informatique et Automatique - Domaine de
Voluceau - Rocquencourt - B.P. 105 - 78153 - LE CHESNAY Cedex - FRANCE''.

INRIA holds all ownership rights relating to SCILAB.

SCILAB is registered at ``Agence pour la protection des programmes (APP)''
under the number 93-27-011-00.

\section{Preamble}
SCILAB is currently being developed and INRIA desires that it can be used by
the scientific community in a spirit of friendship and cooperation. To this
end, INRIA has decided to have SCILAB distributed in source code format by
FTP. However, copyright subsists on the part of SCILAB which has been 
developed at INRIA. SCILAB's numeric library makes use of several public
domain or freely - distributed (and copyrighted by other organizations)
programs.

\section{Extent of the rights granted by INRIA to the users of SCILAB}
INRIA freely grants the right to use, modify and integrate SCILAB in another
program provided that :

-this copyright notice appears in all copies

-acknowledgement of receiving SCILAB is made by returning the agreement below

\noindent
Any modification to SCILAB must be brought to the notice of any other user of
the modified SCILAB.

\section{Reproduction of SCILAB}
INRIA grants any user of SCILAB the right to reproduce it in order to
circulate it in accordance with the same purposes and conditions as those
defined above.

Any copy of SCILAB and/or relevant documentation must include reference to the
ownership of INRIA and the present file.

The user undertakes not to carry out any paying distribution of SCILAB.

\ 

\ 

\ 

\ 


\section{Agreement}
INRIA grants to

\noindent
user name :\\
address	 :\\

\vspace{1.5cm}

\noindent
a license to use, copy  and modify SCILAB under the terms expressed
above. SCILAB is provided `as is', without express or implied warranty.


INRIA will not be responsible for the correction of any bugs or other 
deficiencies . In no event INRIA shall be liable for any damage of any 
kind, including special, indirect or consequent damages, arising in 
connection with the use of SCILAB.\\

\vskip0.6cm
\noindent
Return address :

Scilab/Meta2 - INRIA Bat. 12

Rocquencourt - B.P. 105

78153 LE CHESNAY Cedex

FRANCE

\section{Email address}
email : Scilab@inria.fr

\vskip0.4cm
\noindent
Date:\qquad \qquad \qquad \qquad \qquad \qquad \qquad Signature:
\end{document}
